---
title: "Lab 3: Time Series Explorer - Data Analysis"
author: "Roshan Adhikari"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(12345)
```

## 1. White Noise and Random Walk

A white noise process consists of uncorrelated random variables with zero mean and constant variance. A random walk is formed by taking the cumulative sum of white noise.

```{r white-noise-random-walk}
# Simulate White Noise
n <- 400
white_noise <- rnorm(n, mean = 0, sd = 1)

# Simulate Random Walk
random_walk <- cumsum(white_noise)

# Plotting
par(mfrow = c(2, 1))
plot(white_noise, type = "l", col = "steelblue", main = "White Noise", ylab = "Y(t)", xlab = "Time")
plot(random_walk, type = "l", col = "firebrick", main = "Random Walk", ylab = "Y(t)", xlab = "Time")
```

## 2. Moving Average: MA(1) Process

An MA(1) process models the current value as a linear combination of the current and previous white noise error terms. The ACF of an MA(1) model drops off sharply after lag 1.

```{r ma1-process}
# Simulate MA(1) with theta = 0.8
ma1_sim <- arima.sim(model = list(ma = 0.8), n = n)

par(mfrow = c(1, 2))
plot(ma1_sim, type = "l", col = "darkorange", main = "MA(1) Series (theta = 0.8)", ylab = "Y(t)", xlab = "Time")
acf(ma1_sim, main = "Sample ACF of MA(1)")
```

## 3. Autoregressive: AR(1) Process

An AR(1) process models the current value based on the immediate past value. For an AR(1) model, the ACF decays exponentially, while the PACF cuts off after lag 1.

```{r ar1-process}
# Simulate AR(1) with phi = 0.8
ar1_sim <- arima.sim(model = list(ar = 0.8), n = n)

par(mfrow = c(1, 3))
plot(ar1_sim, type = "l", col = "seagreen", main = "AR(1) Series (phi = 0.8)", ylab = "Y(t)", xlab = "Time")
acf(ar1_sim, main = "ACF of AR(1)")
pacf(ar1_sim, main = "PACF of AR(1)")
```

## 4. Trend and Differencing

A deterministic linear trend renders a series non-stationary. By taking the first difference of the series, we can remove the linear trend and stabilize the mean.

```{r trend-differencing}
# Create a series with a linear trend
time_index <- 1:n
trend_series <- 0.05 * time_index + white_noise

# Apply First Differencing
differenced_series <- diff(trend_series)

par(mfrow = c(2, 2))
plot(trend_series, type = "l", col = "purple", main = "Series with Linear Trend", ylab = "Y(t)", xlab = "Time")
acf(trend_series, main = "ACF (Trend)", lag.max = 30)

plot(differenced_series, type = "l", col = "darkcyan", main = "First Differenced Series", ylab = "Diff(Y(t))", xlab = "Time")
acf(differenced_series, main = "ACF (Differenced)", lag.max = 30)
```

## 5. Seasonality

Seasonal patterns introduce cyclical fluctuations at regular intervals. The ACF of a seasonal series will show significant spikes at multiples of the seasonal period (e.g., lag 12 for monthly data).

```{r seasonality}
# Simulate Seasonality (Period = 12)
seasonal_component <- 5 * sin(2 * pi * time_index / 12)
seasonal_series <- seasonal_component + white_noise

par(mfrow = c(1, 2))
plot(seasonal_series[1:100], type = "l", col = "darkred", main = "Seasonal Series (Subset)", ylab = "Y(t)", xlab = "Time")
acf(seasonal_series, main = "ACF of Seasonal Series", lag.max = 40)
```
